### 2025-01-06 reviewer-response.md

*Near line 154, it is important to give an overview of your solution
strategy and, in particular, why clustering is used first.  Clustering is
a heavily overloaded word.  I would have chosen a more neutral term
instead of "cluster".*

We do define our clustering approach, and moreover, we build on prior work (some ours, some others) that relies on clustering. We respectfully would prefer to stay with the language we have here.

*On line 155, there is the parenthetical comment about "a smaller sample of
the points".  What does that mean precisely and where does it occur later
in the manuscript?*

We clarified this by changing the language about "a smaller sample of the points" to "a random subsample of $\sqrt{|C|}$ points."

@@

*On line 156, there is the claim "so it is a real data point".  What is a
real data point?  Please explain.*

We removed the language "real data point" and clarified that the center of a cluster is the point which is its geometric median.

@@

*The fonts in Algorithms 2.2 through 2.7 are simply too small.  Make them
the font size of the remainder of the text.*

we will make this more legible (NMD)

*The numbers in Table 2 are too small.  Refactor into multiple tables
perhaps.*

We have broken this table up into four tables (one per dataset) which we believe helps readability.

@@

*In line 642, why is "harmonic" italicized?*

This has been fixed.

*The References section needs a through going over for proper formatting
and capitalization.  For example, check the title capitalization in [2],
[4], and [8].  Check the journal name capitalization in [9] and [21],
among others.*

We have capitalized titles as published (confirmed for all of the references mentioned). We have corrected the journal capitalization where appropriate.

### 2025-01-06 sections/1-introduction.tex
addressing minor comments

Researchers are collecting data at an unprecedented scale.
In many fields, datasets grow exponentially, and this increase in the rate of data collection outpaces improvements in computing performance as predicted by Moore's Law~\cite{kahn2011future}.
This indicates that the performance of computing systems will not keep pace with the growth of data.
Often dubbed ``the Big Data explosion,'' this phenomenon has created a need for better algorithms to analyze large datasets.

@@

Given some measure of similarity between data points, there are two common definitions of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
$k$-NN search aims to find the $k$ most similar points to a query, while $\rho$-NN search aims to find all points within a similarity threshold $\rho$ of a query.
Previous works have used the term \textit{approximate} search to refer to $\rho$-NN search, but in this paper, we reserve the term \textit{approximate} for algorithms which do not exhibit perfect recall.
In contrast, an \textit{exact} search algorithm exhibits perfect recall.

$k$-NN search is one of the most ubiquitous classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}.
Na\"{i}ve implementations of $k$-NN search, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets.
While fast algorithms for $k$-NN search on large datasets exist, they are often approximate~\cite{gao2023high}, and while approximate search may be sufficient for some applications, the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.
For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is evidence that distance functions which do not obey the triangle inequality, such as cosine distance, perform poorly for $k$-NN search in biomedical settings~\cite{hu2016distance};
this suggests that approximate $k$-NN search could perform poorly in such contexts.

@@

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy Scaling Search), a set of three novel algorithms for exact $k$-NN search.
We also present improvements to the clustering and $\rho$-NN search algorithms in CHESS~\cite{ishaq2019clustered}, as well as improved genericity across distance functions.
We provide a comparison to several current algorithms for similarity search; namely, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We further benchmark CAKES on a large genomic dataset, the SILVA 18S dataset~\cite{10.1093/nar/gks1219}, using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and a radio frequency dataset, RadioML~\cite{oshea2018radioml}, using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
Finally, we compare real-world datasets to an artificially-generated dataset obeying simple statistical properties.

@@

InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} clusters data into high-dimensional Voronoi cells, and whichever cell a query point falls into is then searched exhaustively, similarly to an early precursor to our work~\cite{yu2015entropy}.
The number of cells used is governed by the $n_{list}$ parameter.
Increasing this parameter decreases the number of points being exhaustively searched, so it improves speed at the cost of accuracy.
To mitigate accuracy issues caused by a query point falling near a cell boundary, the algorithm has a tunable parameter $n_{probe}$, which specifies the number of additional adjacent or nearby cells to search.

@@

\subsection{Entropy Scaling Search}
\label{sec:intoduction:entropy-scaling-search}

>>>>

Researchers are collecting data at an unprecedented rate.
In many fields, datasets grow exponentially, and this increase in the rate of data collection outpaces improvements in computing performance as predicted by Moore's Law~\cite{kahn2011future}.
This indicates that the performance of computing systems will not keep pace with the growth of data.
Often dubbed ``the Big Data explosion,'' this phenomenon has created a need for better algorithms to analyze large datasets.

@@

Given some measure of similarity between data points, there are two common definitions of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
$k$-NN search aims to find the $k$ most similar points to a query, while $\rho$-NN search aims to find all points within a similarity threshold $\rho$ of a query.
Previous works have used the term \textit{approximate} search to refer to $\rho$-NN search, but in this paper, we reserve the term \textit{approximate} for algorithms that do not exhibit perfect recall.
In contrast, an \textit{exact} search algorithm exhibits perfect recall.

$k$-NN search is one of the most ubiquitous classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}.
Na\"{i}ve implementations of $k$-NN search, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets.
While fast algorithms for $k$-NN search on large datasets exist, they are often approximate~\cite{gao2023high}, and while approximate search may be sufficient for some applications, the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.
For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is evidence that distance functions that do not obey the triangle inequality, such as cosine distance, perform poorly for $k$-NN search in biomedical settings~\cite{hu2016distance};
this suggests that approximate $k$-NN search could perform poorly in such contexts.

@@

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy-Scaling Search), a set of three novel algorithms for exact $k$-NN search.
We also present improvements to the clustering and $\rho$-NN search algorithms in CHESS~\cite{ishaq2019clustered}, as well as improved genericity across distance functions.
We provide a comparison to several current algorithms for similarity search; namely, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We further benchmark CAKES on a large genomic dataset, the SILVA 18S dataset~\cite{10.1093/nar/gks1219}, using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and a radio frequency dataset, RadioML~\cite{oshea2018radioml}, using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
Finally, we compare real-world datasets to an artificially-generated dataset obeying simple statistical properties.

@@

InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} clusters data into high-dimensional Voronoi cells, and whichever cell a query point falls into is then searched exhaustively, similarly to an early precursor to our work~\cite{yu2015entropy}.
The number of cells used is governed by the $n_{list}$ parameter.
Increasing this parameter decreases the number of points being exhaustively searched, so it improves speed at the cost of accuracy.
To mitigate accuracy issues caused by a query point falling near a cell boundary, the algorithm has a tunable parameter $n_{probe}$ that specifies the number of additional adjacent or nearby cells to search.

@@

\subsection{Entropy-Scaling Search}
\label{sec:intoduction:entropy-scaling-search}

### 2025-01-06 sections/2-methods.tex
addressing minor comments

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains genomic sequences with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but potentially highly variable at larger length scales, i.e.,\,when $r$ is large.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points in the cluster, i.e.,\,it is the point which minimizes the sum of distances to all other points in the cluster.
In cases where the cardinality of the cluster is large, we take a random subsample of $\sqrt{|C|}$ points and compute the geometric median of that subsample~\cite{ishaq2019clustered}.
The center, therefore, is one of the points in the cluster and is used as a representative of the cluster.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e.,\,$C(c, r) \subset B(c, r)$.

@@

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in this sample.
Using these distances, we compute the \textit{geometric median} of this sample; in other words, we find the point which minimizes the sum of distances to all other points in the sample.
This geometric median is the center of $C$.

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e.,\,the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.
Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

@@

We also note here that this $\mathcal{O}(n \log n)$ complexity assumes that the tree is balanced.
This can realistically only happen when the data are uniformly distributed in the embedding space.
This represents the worst-case scenario for subsequent search algorithms because the LFD of the dataset will approach or equal the embedding dimension.
For real-world datasets with low LFD, we expect the tree to be highly unbalanced, reflecting the topological structure of the data.
When building trees for such datasets, their depths would be larger than $\log n$, which would seem to increase the cost of building
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of the tree would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

@@

% We conduct $\rho$-NN search as described in~\cite{ishaq2019clustered}, but with the following improvement:
% when a cluster overlaps with the query ball, instead of always proceeding to search both of its children, we proceed only with those children which might contain points in the query ball.

@@

% To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
% Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points which are no more than a distance $\rho$ from the query.
% The asymptotic complexity of $\rho$-NN is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn-search-complexity}.

@@

In these algorithms, we use $H$, for \textit{hits}, to refer to the data structure which stores the closest points to the query found so far, and $Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the $k$ nearest neighbors.
The two Sieve algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

@@

We find it useful to start by discussing the \textit{tree-search} and \textit{leaf-search} stages separately.
Tree-search refers to the process of identifying clusters which have overlap with the query ball, or in other words, clusters which might contain one of the $k$ nearest neighbors.
Leaf-search refers to the process of identifying the $k$ nearest neighbors among the points in the clusters identified by tree-search.

@@

To determine the asymptotic complexity of leaf-search, we must estimate $\sum_{C \in Q} |C|$, the total cardinality of the clusters returned by tree-search.
Since we must examine every point in each such cluster, time complexity of leaf-search is linear in $\sum_{C \in Q} |C|$.
Let $\rho_k$ be the distance from the query to the $k^{th}$ nearest neighbor.
Then, we see that $Q$ is expected to be the set of clusters which overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.

@@

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level, and $m$ be a user-specified integer multiplier.
For each datum $x \in X$, we create $m - 1$ new data points within a distance $\epsilon \cdot ||x||$ of $x$ where $||x||$ is the Euclidean distance from $x$ to the origin.
We construct a random vector $r$ of $d$ dimensions in the hyper-sphere of radius $\epsilon$ centered at the origin.
We then add $r$ to $x$ to get a new point $x'$.
Since $||r|| \leq \epsilon$, we have that $||x - x'|| \leq \epsilon$ (i.e.,\,$x'$ is within a distance $\epsilon$ of $x$).
This produces a new dataset $X'$ with $|X'| = m \cdot |X|$.
This augmentation process does not add to the overall topological structure of the dataset, but it does increase its cardinality by a factor of $m$.
This allows us to isolate the effect of cardinality on search performance from that of other factors such as dimensionality, choice of metric, or the topological structure of the dataset.

>>>>

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset that contains genomic sequences with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but potentially highly variable at larger length scales, i.e.,\,when $r$ is large.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points in the cluster, i.e.,\,it is the point that minimizes the sum of distances to all other points in the cluster.
In cases where the cardinality of the cluster is large, we take a random subsample of $\sqrt{|C|}$ points and compute the geometric median of that subsample~\cite{ishaq2019clustered}.
The center, therefore, is one of the points in the cluster and is used as a representative of the cluster.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e.,\,$C(c, r) \subset B(c, r)$.

@@

For a cluster $C$ with $|C|$ points, we begin by taking a random subsample $S$ of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in $S$.
Using these distances, we compute the \textit{geometric median} of $S$; in other words, we find the point that minimizes the sum of distances to all other points in $S$.
We define the \textit{center} of $C$ to be this geometric median.

The \textit{radius} of $C$ is the maximum distance from the center to any other point in $C$.
The point that is responsible for that radius (i.e.,\,the furthest point from the center) is designated the \textit{left pole} and the point that is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster that are closer to the left pole than to the right pole, and the right child contains all points in the cluster that are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points that are equidistant from the two poles.
Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

@@

We also note here that this $\mathcal{O}(n \log n)$ complexity assumes that the tree is balanced.
This can realistically only happen when the data are uniformly distributed in the embedding space.
This represents the worst-case scenario for subsequent search algorithms because the LFD of the dataset will approach or equal the embedding dimension.
For real-world datasets with low LFD, we expect the tree to be highly unbalanced, reflecting the topological structure of the data.
When building trees for such datasets, their depths would be larger than $\log n$, which would seem to increase the cost of tree-building.
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of the tree would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

@@

% We conduct $\rho$-NN search as described in~\cite{ishaq2019clustered}, but with the following improvement:
% when a cluster overlaps with the query ball, instead of always proceeding to search both of its children, we proceed only with those children that might contain points in the query ball.

@@

% To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters that overlap with the query ball or any clusters which lie entirely within the query ball.
% Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points that are no more than a distance $\rho$ from the query.
% The asymptotic complexity of $\rho$-NN is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn-search-complexity}.

@@

In these algorithms, we use $H$, for \textit{hits}, to refer to the data structure that stores the closest points to the query found so far, and $Q$ to refer to the data structure that stores the clusters and points that are still in contention for being one of the $k$ nearest neighbors.
The two Sieve algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

@@

We find it useful to start by discussing the \textit{tree-search} and \textit{leaf-search} stages separately.
Tree-search refers to the process of identifying clusters that overlap with the query ball, or in other words, clusters that might contain one of the $k$ nearest neighbors.
Leaf-search refers to the process of identifying the $k$ nearest neighbors among the points in the clusters identified by tree-search.

@@

To determine the asymptotic complexity of leaf-search, we must estimate $\sum_{C \in Q} |C|$, the total cardinality of the clusters returned by tree-search.
Since we must examine every point in each such cluster, time complexity of leaf-search is linear in $\sum_{C \in Q} |C|$.
Let $\rho_k$ be the distance from the query to the $k^{th}$ nearest neighbor.
Then, we see that $Q$ is expected to be the set of clusters that overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + 2\hat{r}$, where $\hat{r}$ is the mean radius of the clusters in $Q$.

@@

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level, and $m$ be a user-specified integer multiplier.
For each datum $\mathbf{x} \in X$, we create $m - 1$ new data points within a distance $\epsilon$ of $\mathbf{x}$.
We construct a random vector $\mathbf{r}$ of $d$ dimensions in the hyper-sphere of radius $\epsilon$ centered at the origin.
We then add $\mathbf{r}$ to $\mathbf{x}$ to get a new point $\mathbf{x}'$.
Since $||\mathbf{r}|| \leq \epsilon$, we have that $||\mathbf{x} - \mathbf{x}'|| \leq \epsilon$ (i.e.,\,$\mathbf{x}'$ is within a distance $\epsilon$ of $\mathbf{x}$).
This produces a new dataset $X'$ with $|X'| = m \cdot |X|$.
This augmentation process does not add to the overall topological structure of the dataset, but it does increase its cardinality by a factor of $m$.
This allows us to isolate the effect of cardinality on search performance from that of other factors such as dimensionality, choice of metric, or the topological structure of the dataset.
